---
title: "DESEQ_analysis"
author: "Sascha Maschmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# setwd("/media/molgenuser/DATA/Sascha_analyses/HPR4_toxo_srna_pipeline/scripts/R")
plot_dir <- "plots/paper_plots" # "/run/user/1000/gvfs/smb-share:server=hubiombp12c.cms.hu-berlin.de,share=biomolgenetik/Lab4/Sabrina/HPR4/paper/plots_Sascha"
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

packages <- c("plotly", "tidyverse", "ggrepel", "circlize")

installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)

if (length(to_install) > 0) {
  message("Installing missing packages: ", paste(to_install, collapse = ", "))
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

library(plotly)
library(tidyverse)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(edgeR)
library(ggsignif)
```

# Analysis of mito reads

## Mapping percentages
Investigation of relative amounts of reads mapping to the pseudo-genome.  
Decreases with treatment.

```{r plot mapping percentages, fig.asp=9/19, fig.width=12}
theme_clean_axes <- function(base_size = 11, base_family = "") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      # Remove background
      panel.background = element_blank(),
      plot.background = element_blank(),
      
      # Remove grid lines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # Axis lines: show only x and y in bold
      axis.line = element_line(colour = "black", linewidth = 1.1),
      
      # Remove axis ticks if you want it even cleaner (optional)
      # axis.ticks = element_blank(),
      
      # Keep axis text and titles as usual
      # axis.title = element_text(face = "bold"),
      axis.text = element_text()
    )
}

pairwise_logit_ttest <- function(df,
                                 cond_col,
                                 value_col,
                                 eps = 1e-4,
                                 p_adjust_method = "BH") {
  # Extract columns
  cond  <- df[[cond_col]]
  vals  <- df[[value_col]]
  
  if (!is.numeric(vals)) {
    stop("value_col must be numeric (percentages or proportions).")
  }
  
  # Auto-detect: if values > 1, treat as percentages and scale to [0,1]
  prop <- ifelse(vals > 1, vals / 100, vals)
  
  # Clamp to (0,1) to avoid logit problems at 0 and 1
  prop <- pmin(pmax(prop, eps), 1 - eps)
  
  logit <- function(p) log(p / (1 - p))
  lvals <- logit(prop)
  
  conditions <- unique(cond)
  combos <- combn(conditions, 2, simplify = FALSE)
  
  res_list <- lapply(combos, function(pair) {
    idx1 <- cond == pair[1]
    idx2 <- cond == pair[2]
    
    x_log <- lvals[idx1]
    y_log <- lvals[idx2]
    
    tt <- t.test(x_log, y_log)  # Welch t-test on logit-scale
    
    # Back to original scale for reporting (use vals, not prop)
    x_vals <- vals[idx1]
    y_vals <- vals[idx2]
    
    data.frame(
      condition1 = as.character(pair[1]),
      condition2 = as.character(pair[2]),
      n1         = sum(idx1),
      n2         = sum(idx2),
      mean1      = mean(x_vals),
      mean2      = mean(y_vals),
      diff       = mean(x_vals) - mean(y_vals),
      t          = unname(tt$statistic),
      df         = unname(tt$parameter),
      p_value    = tt$p.value,
      stringsAsFactors = FALSE
    )
  })
  
  res <- do.call(rbind, res_list)
  res$p_adj <- p.adjust(res$p_value, method = p_adjust_method)
  
  # Sort by raw p-value
  res[order(res$p_value), ]
}



mapping_perc <- read.table("../../out/mapping_percentages.txt", header = T) %>% 
  mutate(con = base::substr(sample, 1,3))


res <- pairwise_logit_ttest(mapping_perc, "con", "percent_mapped")

res_plot <- filter(res, condition1 == "pr0") %>% 
  mutate(annotation = case_when(p_adj <= 0.001 ~ "***",
                                p_adj <= 0.01 ~ "**",
                                p_adj <= 0.05 ~ "*",
                                TRUE ~ "n.s.")) %>% 
  filter(annotation != "n.s.")


mapping_perc_sum <- read.table("../../out/mapping_percentages.txt", header = T) %>% 
  mutate(con = base::substr(sample, 1,3)) %>% 
  group_by(con) %>% 
  summarise(mean = mean(percent_mapped),
            SEM = sd(percent_mapped) / sqrt(n()),
            min = min(percent_mapped),
            max = max(percent_mapped)) %>% 
  mutate(col = ifelse(con == "par", "#3A3A3A", "#6C8880"))


# top of error bar per condition
y_tops <- mapping_perc_sum %>%
  mutate(y_top = mean + SEM) %>%
  select(con, y_top)

# build a df for ggsignif with xmin, xmax, annotation, y_position
offset <- 2  # vertical spacing in percentage points

signif_df <- res_plot %>%
  # join y_top for pr0 and the other condition
  left_join(y_tops, by = c("condition1" = "con")) %>%
  rename(y1 = y_top) %>%
  left_join(y_tops, by = c("condition2" = "con")) %>%
  rename(y2 = y_top) %>%
  mutate(
    xmin       = condition1,
    xmax       = condition2,
    y_base     = pmax(y1, y2),
    # stack brackets: 1st a bit above, 2nd higher, etc.
    y_position = y_base + offset * (row_number())
  ) %>%
  select(xmin, xmax, annotation, y_position)

ggplot(mapping_perc_sum,
       aes(x = con, y = mean)) +
  geom_point(aes(color = col), size = 0.25) +
  geom_linerange(aes(ymin = mean - SEM,
                     ymax = mean + SEM,
                     color = col),
                 linewidth = 0.05) +
  scale_color_identity()+
  geom_signif(
    data   = signif_df,
    aes(xmin = xmin,
        xmax = xmax,
        annotations = annotation,
        y_position = y_position),
    manual    = TRUE,
    tip_length = 0,
    textsize   = 1.5,
    size = 0.05
  ) +
  theme_diss() +
  scale_x_discrete(labels = c(
  "par" = "Parental \nDay 0",
  "pr0" = "rHPR4-FLAG \nDay 0",
  "pr1" = "rHPR4-FLAG \nDay 1",
  "pr2" = "rHPR4-FLAG \nDay 2",
  "pr3" = "rHPR4-FLAG \nDay 3"
)) +
  labs(
    x     = "",
    y     = "Fraction of Total Reads Mapped to Mito Genome [%]",
  )

ggsave(filename = "~/diss_plots/mrsic_mapping_percentages.svg",
       device = "svg",
       width = 50,
       height = 40,
       units = "mm",
       )

```

## Normalisation
Using A set of stable genes encoded in the nuclear or apicoplast genome, mito mapped reads are normalized.

```{r load data}
countFiles <- list.files("../../out/counts/mito/shortFeatures/sense",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

nSample <- length(countFiles)

absoluteNormalizators <- base::readRDS("intermediates/absolute_norm_values.RDS")

coldata <- read.csv("../../samples.csv")

mitoCounts <- read.delim(countFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in countFiles[2:length(countFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  mitoCounts <- left_join(mitoCounts, counts, by = "Geneid")
}

longFiles <- list.files("../../out/counts/mito/longFeatures",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

# Filter out all genes with less than 20 reads in all samples
cleanShortCounts <- mitoCounts %>% 
  filter(if_all(-Geneid, ~ .x >= 20)) %>% 
  # Remove 'rRNA' from Geneids
  mutate(Geneid = base::gsub(".rRNA", "", Geneid)) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

longCounts <- read.delim(longFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in longFiles[2:length(longFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  longCounts <- left_join(longCounts, counts, by = "Geneid")
}

cleanLongCounts <- longCounts %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

cleanCounts <- rbind(cleanShortCounts, cleanLongCounts)
```

```{r plot example genes, fig.asp=9/19, fig.width=12}
example_genes = c("SSUB", "RNA10_F", "RNA10_1", "RNA10_2", "RNA33_F", "RNA33_1", "RNA33_2")


example_plot <- cleanCounts %>% 
  filter(Geneid %in% example_genes) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample_name") %>% 
  mutate(con = base::substr(sample_name, 1,3))

ggplot(example_plot,
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  theme_bw() +
  scale_y_log10() +
  labs(x = "Condition",
       y = expression(log[10]("Absolute Reads")),
       title = "Raw Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")


```
```{r Normalize}
# mitoCounts: matrix of mito features (rows) x samples (columns), raw counts
mitoCounts <- cleanCounts %>%
  mutate(Geneid = paste0(Geneid, "XX")) %>%
  column_to_rownames("Geneid") 

# group: factor of conditions for each sample (length = ncol(mitoCounts))
group <- factor(base::substr(colnames(mitoCounts), 1,3), levels = unique(base::substr(colnames(mitoCounts), 1,3)))

# anchor A_j:   numeric vector of length ncol(mitoCounts)
#               IMPORTANT: A_j must exclude mitochondrial reads; >0 in every sample
Aj <- absoluteNormalizators$value


# 1) Prepare DGEList with offset-only normalization (NO TMM)
y  <- DGEList(mitoCounts)

y$samples$lib.size <- Aj
y$samples$group <- group

# 2) Design with all groups (no intercept), pr_0 will be the baseline in contrasts
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

# 3) Dispersion + fit (QL pipeline)
y  <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# 4) Define contrasts: (first) - (pr_0)
CT <- makeContrasts(
  par_pr0 = par - pr0,
  pr0_pr1 = pr1    - pr0,   # logFC > 0 => pr_1 > pr_0
  pr0_pr2 = pr2    - pr0,
  pr0_pr3 = pr3    - pr0,
  levels = design
)

# 5) Run tests and collect results
run_one <- function(fit, contr_name) {
  tt <- glmQLFTest(fit, contrast = CT[, contr_name])
  out <- topTags(tt, n = Inf)$table
  out$contrast <- contr_name
  out
}

res_list <- lapply(colnames(CT), function(nm) run_one(fit, nm))
res <- do.call(rbind, res_list)

# res has logFC (relative to the anchor), PValue, FDR for each feature/bin and contrast
split(res, res$contrast) |>
  lapply(function(df) df[order(df$FDR), ]) -> by_contrast

res_clean <- res %>% 
  rownames_to_column("Geneid") %>% 
  mutate(Geneid = sub("XX.*", "", Geneid))
```

```{r plot normalized reads, fig.asp=9/19, fig.width=12}

normReads <- data.frame(norm_counts <- cpm(y, normalized.lib.sizes = TRUE)) %>%
  rownames_to_column("Geneid") %>%
  mutate(Geneid = sub("XX", "", Geneid)) %>%
  pivot_longer(cols = 1:nSample+1,
               names_to = "sample_name",
               values_to = "value") %>% 
  mutate(con = base::substr(sample_name, 1,3))

norm_reads_sum <- normReads %>% 
  group_by(sample_name) %>% 
  summarise(val = sum(value)) %>% 
  mutate(con = base::substr(sample_name, 1,3),
         norm = "modified") %>% 
  bind_rows(norm_reads_sum_base) %>% 
  ungroup() %>% 
  dplyr::group_by(norm) %>% 
  mutate(val_scale = (val- min(val)) /(max(val)-min(val))) %>% 
  ungroup() %>% 
  group_by(con, norm) %>% 
  mutate(val_mean = mean(val_scale),
         val_sem = sd(val_scale) / sqrt(n()))
  

ggplot(filter(normReads, Geneid %in% example_genes),
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Condition",
       y = expression(log[10]("Normalized Reads")),
       title = "Normalized Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")

saveRDS(norm_reads_sum,
        "intermediates/normalisation_comp.RDS")

ggplot(norm_reads_sum,
       aes(x = con,
           y = val_mean)) +
  geom_point(size = 0.25) +
  geom_linerange(aes(ymin = val_mean - val_sem,
                     ymax = val_mean + val_sem),
                 linewidth = 0.05) +
  theme_bw() +
  facet_wrap(~norm) +
  labs(y = "Normalized CPM") +
  theme_diss() +
    theme(axis.title.x = element_blank())

ggsave(filename = "~/diss_plots/mrsic_norm_comp_total.svg",
       device = "svg",
       width = 100,
       height = 40,
       units = "mm",
       )
```

```{r PCA, fig.asp=9/19, fig.width=12}
PCA_Reads <- cpm(y, log = TRUE, normalized.lib.sizes = TRUE)
PCA_reads <- cleanCounts %>% as.matrix()


pca <- prcomp(t(PCA_Reads), scale. = TRUE)
ve <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
pc <- pca

autoplot <- function(pc, col) {
  df <- data.frame(pc$x[,1:2], group = base::substr(rownames(pc$x), 1,3))
  ggplot(df, aes(PC1, PC2, shape = group)) +
    geom_point(size = 2) +
    labs(x = paste0("PC1 (", ve[1], "%)"),
         y = paste0("PC2 (", ve[2], "%)"),
         title = "PCA of Normalized Mito Reads") +
    theme_bw()
}

p1 <- autoplot(pca, "con")
p1

```

## DEG Analysis
Based on the normalized reads, log2fold changes and statistical significance are calculated.  
Below results are shown in various ways.

```{r Heatmap all Genes, fig.height=9, fig.width=3.5}

lfc_matrix <- res_clean %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  column_to_rownames("Geneid") %>% 
  as.matrix()

col_fun1 = colorRamp2(c(ceiling(max(lfc_matrix, na.rm = TRUE)), 0, floor(min(lfc_matrix, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_all <- Heatmap(lfc_matrix,
        column_title = "LFCs mito rRNA fragments",
        cluster_columns = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC",
        )
plot(heatmap_all)
```

```{r}
relevant_vals <- cleanCounts %>% 
  select(Geneid, par1, par2, par3, pr0_1, pr0_2, pr0_3)

uptick_expl <- volcano %>% 
  filter(contrast == "par_pr0") %>% 
  select(Geneid, logFC, contrast) %>% 
  left_join(relevant_vals) %>% 
  pivot_longer(cols = 4:9) %>% 
  mutate(trend = as.character(sign(logFC)))

ggplot(uptick_expl,
       aes(x = trend,
           y = value)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.6) +
  scale_y_log10() +
  theme_bw() 
```


Based on unfiltered log2fold changes, the general trend continues: over all, mito rRNA transcription presumably is dampened by the suppression of HPR4 transcription.

```{r filtered Heatmap, fig.height=7.5, fig.width=3.5}
p_cutoff <- 0.05
fdr_cutoff <- 0.05

lfc_matrix_minimal <- res_clean %>% 
  mutate(logFC = ifelse(PValue <= p_cutoff & FDR <= fdr_cutoff, logFC, NA)) %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  filter(if_any(-Geneid, ~!is.na(.x))) %>% 
  select_if(~sum(!is.na(.)) > 0) %>% 
  column_to_rownames("Geneid") %>% 
  arrange(pr0_pr2, pr0_pr3) %>%
  as.matrix()

col_fun2 = colorRamp2(c(2, 0, floor(min(lfc_matrix_minimal, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_sig <- Heatmap(lfc_matrix_minimal,
        column_title = "Significantly changed \n mito rRNA fragments",
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC"
        )
plot(heatmap_sig)
```

When filtering based on adjusted p value (\<= 0.05) only decreases in rRNA fragment expression remain.
, fig.asp=9/19, fig.width=8
```{r Volcano}
units <- read.csv("intermediates/unit_assignment.csv")%>%
  group_by(Geneid) %>%
  summarise(
    unit = if (n_distinct(unit) > 1) "both" else unique(unit)
  ) %>%
  ungroup()

GOI <- c("RNA33_1", "RNA33_F", "RNA42")
# In Case we want SSUB and RNA 7 in:
# , "RNA7", "SSUB"


volcano <- res_clean %>% 
  select(-logCPM, -`F`) %>% 
  filter(!(Geneid %in% c("cob", "coxI", "coxIII"))) %>% 
  mutate(y_var = -log10(FDR),
         label = ifelse(
           # PValue <= p_cutoff & # Comment out if need to show SSUB
           #   FDR <= fdr_cutoff &
             Geneid %in% GOI, Geneid, ""),
         Geneid2 = Geneid,
         Geneid = base::gsub("_F", "", Geneid)
  ) %>% 
  left_join(units, by = "Geneid") %>% 
  mutate(Geneid = Geneid2,
         unit = case_when(!is.na(unit) ~ unit,
                          grepl("LSU", Geneid) ~ "LSU",
                          grepl("SSU", Geneid) ~ "SSU",
                          is.na(unit) ~ "unassigned"
                          )) %>% 
  mutate(color = ifelse(FDR <= fdr_cutoff & abs(logFC) >= 1, unit, "grey"),
         color = case_when(
           color == "LSU" ~ "#ff33ff",
           color == "SSU" ~ "#cc00ff",
           color == "both" ~ "#9900ff",
           color == "unassigned" ~ "black",
           TRUE ~ "grey"
         ),
         shape = case_when(
           unit == "LSU" ~ 3,
           unit == "SSU" ~ 4,
           unit == "both" ~ 8,
           unit == "unassigned" ~ 0
         )) %>% 
  select(-Geneid2)
# Common scale for all three plots
ribosomal_fill_scale <- scale_fill_identity(
  name   = "Ribosomal Unit",
  guide  = "legend",
  breaks = c("#ff33ff", "#cc00ff", "#9900ff", "black"),
  labels = c("LSU", "SSU", "both", "unassigned")
)

# Helper to build one volcano plot for a given contrast + title
make_volcano <- function(contrast_id, title_text) {
  ggplot(
    data = filter(volcano, contrast == contrast_id),
    aes(x = logFC, y = y_var)
  ) +
    geom_text_repel(
      aes(label = label),
      segment.color      = "black",
      min.segment.length = 0.1,
      max.overlaps       = 20,
      size               = 3,
      force              = 8,
      force_pull         = 0.2
    ) +
    geom_point(
      aes(fill = color),
      shape = 21,
      size  = 2
    ) +
    ribosomal_fill_scale +
    geom_hline(
      yintercept = -log10(fdr_cutoff),
      linetype   = "dashed",
      color      = "grey40"
    ) +
    geom_vline(
      xintercept = c(-1, 1),
      linetype   = "dashed",
      color      = "grey40"
    ) +
    labs(
      x     = expression(log[2]("fold change")),
      y     = expression(-log[10](padj)),
      title = title_text
    ) +
    theme_bw()
}

# Now your three plots are just:
volcano_0_1 <- make_volcano("pr0_pr1", "Day 0 vs Day 1")
volcano_0_2 <- make_volcano("pr0_pr2", "Day 0 vs Day 2")
volcano_0_3 <- make_volcano("pr0_pr3", "Day 0 vs Day 3")
volcano_0_0 <- make_volcano("par_pr0", "Day 0 vs Day 0")


plot(volcano_0_1)
plot(volcano_0_2)
plot(volcano_0_3)
plot(volcano_0_0)

ggsave(filename = "volcano_0_1.svg",
       path = plot_dir,
       plot = volcano_0_1,
       device = "svg",
       width = 16,
       height = 11,
       units = "cm",
       )

ggsave(filename = "volcano_0_2.svg",
       path = plot_dir,
       plot = volcano_0_2,
       device = "svg",
       width = 16,
       height = 11,
       units = "cm",
       )

ggsave(filename = "volcano_0_3.svg",
       path = plot_dir,
       plot = volcano_0_3,
       device = "svg",
       width = 16,
       height = 11,
       units = "cm",
       )

```

```{r}
volcano_faceted <- volcano %>%
  filter(contrast %in% c("pr0_pr1", "pr0_pr2", "pr0_pr3")) %>%
  mutate(
    contrast_label = factor(
      contrast,
      levels = c("pr0_pr1",      "pr0_pr2",      "pr0_pr3"),
      labels = c("Day 0 vs Day 1", "Day 0 vs Day 2", "Day 0 vs Day 3")
    )
  )

volcano_facet <- ggplot(
  data = volcano_faceted,
  aes(x = logFC, y = y_var)
) +
  geom_text_repel(
    aes(label = label),
    segment.color      = "black",
    min.segment.length = 0.1,
    point.size = 3,
    max.overlaps       = 40,
    size               = 3,
    force              = 8,
    force_pull         = 0.2
  ) +
  geom_point(
    aes(fill = color),
    shape = 21,
    size  = 3
  ) +
  ribosomal_fill_scale +
  geom_hline(
    yintercept = -log10(fdr_cutoff),
    linetype   = "dashed",
    color      = "grey40"
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype   = "dashed",
    color      = "grey40"
  ) +
  labs(
    x = expression(log[2] * "(fold change)"),
    y = expression("-"*log[10] * "(padj)")
  ) +
  facet_grid(
    rows  = vars(contrast_label),
    cols  = vars(),            # important: avoid column facets
    space  = "free_x",         # ensures spacing matches data
    switch = "x"               # puts x-axis at top of each panel
  ) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0),  # readable facet labels
    legend.position = "right")

print(volcano_facet)

ggsave(plot = volcano_facet,
       path = plot_dir,
       filename = "volcano_facetted.svg",
       device = "svg",
       width = 340/2,
       height = 401/2,
       units = "mm",
       )
```


## Quantification of Subunits affected
The table below highlights the ratio of large and small ribosome subunit fragments.

```{r quantify unit effect}

unitEffect <- volcano %>% 
  select(Geneid, FDR, logFC, contrast, unit) %>% 
  filter(FDR <= fdr_cutoff) %>% 
  mutate(expressed = ifelse(sign(logFC) == -1, "down", "up")) %>% 
  group_by(contrast, expressed) %>% 
  mutate(totalDE = n()) %>% 
  ungroup() %>% 
  group_by(contrast, unit, expressed) %>% 
  summarise(relative = n() / totalDE) %>% 
  distinct()

print(unitEffect)
```

## Compare Fragments per RNA 'identity'
Grey Log2FCs are not significant.

```{r multi fragment cases, fig.asp=9/19, fig.width=12}

deMultiFrag <- res_clean %>% 
  mutate(Geneid2 = Geneid,
         Geneid = base::gsub("_.*", "", Geneid)
  ) %>% 
  select(Geneid, Geneid2, logFC, contrast, FDR) %>% 
  group_by(Geneid)  %>% 
  group_split()

test <- deMultiFrag[[17]] %>% 
  mutate(Geneid2 = ifelse(grepl("_", Geneid2), sub("^[^_]*_", "", Geneid2), "F")) %>% 
  pivot_wider(values_from = c(logFC, FDR), names_from = Geneid2)

untangler <- function(df){
  df_untangled <- df %>% 
    mutate(Geneid2 = ifelse(grepl("_", Geneid2), sub("^[^_]*_", "", Geneid2), "F")) %>% 
    pivot_wider(values_from = c(logFC, FDR), names_from = Geneid2)
  return(df_untangled)
}

fragTable <- lapply(deMultiFrag, untangler) %>%
  bind_rows()

trueFragmented <- fragTable %>% 
  filter(!is.na(logFC_1)) %>% 
  pivot_longer(cols = 3:8,
               names_to = c("stat", "fragment"),
               names_sep = "_") %>% 
  pivot_wider(names_from = "stat",
              values_from = "value") %>% 
  mutate(color = ifelse(FDR <= fdr_cutoff, "black", "grey"))

for (id in unique(trueFragmented$Geneid)){
  plot <- ggplot(
    data = filter(trueFragmented, Geneid == id),
    aes(x = contrast,
        y = logFC, 
        shape = fragment,
        color = color)
    ) +
    scale_color_identity() +
    geom_point(size = 2) +
    theme_bw() +
    labs(x = "",
         y = 'LFC',
         title = id)
  
  plot(plot)
}
```


## Relative number of significantly changed rRNAs

Below two variants of relative number of significantly changed rRNA are plotted, one only respects the full length transcripts, the other all annotated transcripts variants.

```{r changed rRNAs, fig.asp=9/19, fig.width=12}
all_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue")) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = nrow(cleanCounts),
         relativeValues = n/total*100)

relativeChanged_all <-
  ggplot(all_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on All Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))


total_full <- cleanCounts %>% 
  filter(!grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  summarise(n()) %>% 
  as.integer()

fullLength_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff,
         !grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue"),
         Geneid = base::gsub("_F", "", Geneid)) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = total_full,
         relativeValues = n/total*100)

relativeChanged_FL <- 
  ggplot(fullLength_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on Full Length Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))

fullLength_changed_pub <- res_clean %>% 
  filter(FDR <= fdr_cutoff,
         # !grepl("_1", Geneid),
         # !grepl("_2", Geneid),
         # Geneid != "RNA42"
         ) %>% 
  separate(contrast, into = c("cond1", "cond2"), sep = "_") %>% 
  mutate(Geneid = base::gsub("_F", "", Geneid),
         logFC = round(logFC, 3),
         FDR = signif(FDR, 3),
         cond1 = case_when(
           cond1 == "par" ~ "mRSiC-FLAG -ATc",
           cond1 == "pr0" ~ "rmRSiC-FLAG -ATc",
           cond1 == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           cond1 == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           cond1 == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         ),
         cond2 = case_when(
           cond2 == "par" ~ "mRSiC-FLAG -ATc",
           cond2 == "pr0" ~ "rmRSiC-FLAG -ATc",
           cond2 == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           cond2 == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           cond2 == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         )) %>% 
  select(Geneid, cond1, cond2, logFC, FDR) %>% 
  arrange(cond1, cond2, logFC) %>% 
  write_csv(file = paste0(plot_dir, "/TableS2_full"))

fullLength_changed_sum_pub <- fullLength_changed_pub %>% 
  mutate(`effect direction` = ifelse(logFC > 0, "up", "down"),
         Geneid = base::gsub("_F", "", Geneid)) %>% 
  group_by(cond1, cond2, `effect direction`) %>% 
  summarise(n = n()) %>% 
  write_csv(file = paste0(plot_dir, "/TableS2_summarised"))

relativeChanged_all
relativeChanged_FL
```

## PolyA Tail plots

```{r polya tail plots, fig.asp=9/19, fig.width=12}
polyA_files <- list.files("../../out/polyA",
                          pattern = "*_polyA_lengths.tsv$",
                          full.names = T)
polyA_raw <- data_frame()
for (f in polyA_files){
  ph <- read.delim(f, header = F) %>% 
    select(Geneid = V1, polyA = V2, n = V3) %>% 
    mutate(sample = f,
           sample = base::gsub("../../out/polyA/", "", sample),
           sample = base::gsub("_polyA_lengths.tsv", "", sample))
  
  polyA_raw <- rbind(polyA_raw, ph)
}

polyA_summarised <- polyA_raw %>% 
  group_by(Geneid, sample) %>% 
  arrange(polyA) %>% 
  mutate(n_rel = n / sum(n),
         cumsum = cumsum(n_rel),
         n_total = sum(n)) %>% 
  filter(cumsum <= 0.99,
         n_total >= 25) %>% 
  ungroup() %>% 
  mutate(condition = base::substr(sample, 1,3)) %>% 
  group_by(Geneid, polyA, condition) %>% 
  summarise(n_mean = mean(n_rel),
            SEM = sd(n_rel) / sqrt(n())) %>% 
  mutate(condition = case_when(
           condition == "par" ~ "mRSiC-FLAG -ATc",
           condition == "pr0" ~ "rmRSiC-FLAG -ATc",
           condition == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           condition == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           condition == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         ),
         Geneid = ifelse(Geneid == "RNA37/38_F", "RNA37-38_F", Geneid))

saveRDS(polyA_summarised,
        file = "mrsic_polyA.RDS")

for (id in unique(polyA_summarised$Geneid)){
  plot <-
    ggplot(filter(polyA_summarised, Geneid == id),
                 aes(x = polyA,
                     y = n_mean,
                     color = condition,
                     fill = condition,
                     ymin = n_mean - SEM,
                     ymax = n_mean + SEM)) +
    geom_linerange(aes()) +
    geom_line() +
    theme_bw() +
    labs(x = "polyA-tail length",
         y = "mean relative frequency",
         title = id)
  
  png(filename = paste0("plots/polyA/", id, ".png"),
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
  plot(plot)
  dev.off()
  plot(plot)
}

polyA_summarised_pub <- polyA_raw %>% 
  group_by(Geneid, sample) %>% 
  arrange(polyA) %>% 
  mutate(n_rel = n / sum(n),
         cumsum = cumsum(n_rel),
         n_total = sum(n)) %>% 
  filter(cumsum <= 0.99,
         n_total >= 25) %>% 
  ungroup() %>% 
  mutate(condition = base::substr(sample, 1,3),
         Geneid = base::gsub("/", "-", Geneid)) %>% 
  group_by(Geneid, polyA, condition) %>% 
  summarise(`mean %` = round(mean(n_rel)*100, 2),
            SEM = round(sd(n_rel)*100 / sqrt(n()), 2)) %>% 
  arrange(Geneid, condition, polyA) %>% 
  mutate(condition = case_when(
           condition == "par" ~ "mRSiC-FLAG -ATc",
           condition == "pr0" ~ "rmRSiC-FLAG -ATc",
           condition == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           condition == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           condition == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         )) %>% 
  write_csv(file = "/run/user/1000/gvfs/smb-share:server=hubiombp12c.cms.hu-berlin.de,share=biomolgenetik/Lab4/Sabrina/HPR4/paper/tables/TableS3")
```

```{r PolyA publications}
library(ggridges)
genes_of_interest <- c("RNA33_1", "RNA33_2", "RNA33_F", "RNA42")
cutoffs <- polyA_summarised %>%
  dplyr::group_by(Geneid, condition) %>%
  dplyr::summarise(polyA_max = max(polyA), .groups = "drop")


condition_labels <- c(
  "par" = "mRSiC-FLAG -ATc",
  "pr0" = "rmRSiC-FLAG -ATc",
  "pr1" = "rmRSiC-FLAG 1d ATc",
  "pr2" = "rmRSiC-FLAG 2d ATc",
  "pr3" = "rmRSiC-FLAG 3d ATc"
)


# for (g in genes_of_interest) {
#   
#     df_g <- polyA_summarised %>%
#       filter(Geneid == g)
#     max_polyA <- max(df_g$polyA)
# 
#   p_ridgeline <- ggplot(
#   df_g,
#   aes(
#     x      = polyA,
#     y      = condition,
#     height = n_mean,
#     fill   = condition,
#     group  = condition
#   )
# ) +
#   geom_ridgeline(
#     alpha  = 0.8,
#     scale  = 7,
#     size   = 0.3,
#     colour = "grey20"
#   ) +
# scale_y_discrete(
#   labels = condition_labels,
#   expand = expansion(mult = c(0.1, 0.2))   # more y-space
# ) +
#     scale_x_continuous(
#     breaks       = seq(0, max_polyA, by = 5),   # labelled ticks
#     minor_breaks = seq(0, max_polyA, by = 1)    # extra grid lines every 1
#   ) +
#     labs(
#     title = g,
#     x = "polyA-tail length",
#     y = NULL,
#     fill = "Condition"
#   ) +
#   theme_bw() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5),
#     panel.spacing = unit(3, "lines")
#   )
#   plot(p_ridgeline)
#   
#   ggsave(
#     filename = paste0("polyA_", g, ".svg"),
#     plot     = p_ridgeline,
#     path = plot_dir,
#     device = "svg",
#     width = 16,
#     height = 11,
#     units = "cm",
#     )
# }

facetted_polyA <- ggplot(
  polyA_summarised,
  aes(
    x      = polyA,
    y      = n_mean,
    color   = condition
  )
) +
  geom_line() +
scale_x_continuous(
    breaks       = seq(0, 25, by = 5),   # labelled ticks
    minor_breaks = seq(0, 25, by = 1)    # extra grid lines every 1
  ) +
    labs(
    x = "polyA-tail length",
    y = NULL,
    color = "Condition"
  ) +
  facet_wrap(facets = ~ Geneid,) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5),
    panel.spacing = unit(.25, "lines")
  )

facetted_polyA

 ggsave(
    filename = paste0("polyA_facetted.svg"),
    plot     = facetted_polyA,
    path = plot_dir,
    device = "svg",
    width = 40,
    height = 30,
    units = "cm",
    )

```

```{r}
## ------------------------------------------------------------------
## Setup
## ------------------------------------------------------------------
library(dplyr)
library(tidyr)
library(purrr)
library(parallel)
library(tibble)
# ggplot2 only needed if you want to plot later
# library(ggplot2)

## ------------------------------------------------------------------
## 0. Preprocess sample names: condition + technical replicate
## ------------------------------------------------------------------
# sample format: "<condition>-<replicate>", e.g. "pr1-3"
polyA <- polyA_raw %>%
  mutate(
    condition  = substr(sample, 1,3),    # "pr1" part
    replicate  = substr(sample, 4,4),    # "3" part
    tech_group = condition                  # here: tech group = condition
  )

## ------------------------------------------------------------------
## 1. Wasserstein distance from counts (no uncount)
## ------------------------------------------------------------------
wasserstein_counts <- function(n1, n2) {
  p1 <- n1 / sum(n1)
  p2 <- n2 / sum(n2)
  cdf1 <- cumsum(p1)
  cdf2 <- cumsum(p2)
  # step size = 1 nt
  sum(abs(cdf1 - cdf2))
}

## ------------------------------------------------------------------
## 2. Permutation on counts (binomial approximation)
## ------------------------------------------------------------------
# m   : vector of total counts per length (n1 + n2)
# p1  : proportion of molecules in condition 1 (N1 / (N1 + N2))
# n_perm : number of permutations
perm_wasserstein_counts <- function(m, p1, n_perm = 1000) {
  L <- length(m)
  w_perm <- numeric(n_perm)
  
  for (i in seq_len(n_perm)) {
    n1_perm <- rbinom(L, size = m, prob = p1)
    n2_perm <- m - n1_perm
    w_perm[i] <- wasserstein_counts(n1_perm, n2_perm)
  }
  
  w_perm
}

## ------------------------------------------------------------------
## 3. Per-gene test for biological comparisons (condition pairs)
## ------------------------------------------------------------------
# df_gene: subset of polyA for a single Geneid
# requires columns: Geneid, sample, polyA, n, condition
test_polyA_counts <- function(df_gene, n_perm = 1000, min_total = 0) {
  conds <- unique(df_gene$condition)
  
  # need at least two conditions
  if (length(conds) < 2L) {
    return(tibble())
  }
  
  # aggregate counts per condition and polyA length
  agg <- df_gene %>%
    group_by(condition, polyA) %>%
    summarise(n = sum(n), .groups = "drop")
  
  # all condition pairs for this gene
  pairs <- combn(conds, 2, simplify = FALSE)
  
  res_list <- lapply(pairs, function(p) {
    sub <- agg %>% filter(condition %in% p)
    
    # support = all lengths observed for this pair
    lengths <- sort(unique(sub$polyA))
    
    # complete to full length support per condition, fill missing with 0
    wide <- sub %>%
      tidyr::complete(condition, polyA = lengths, fill = list(n = 0)) %>%
      tidyr::pivot_wider(names_from = condition, values_from = n)
    
    n1 <- as.numeric(wide[[p[1]]])
    n2 <- as.numeric(wide[[p[2]]])
    
    total_counts <- sum(n1) + sum(n2)
    if (total_counts < min_total) {
      return(NULL)
    }
    
    # observed Wasserstein effect size
    w_obs <- wasserstein_counts(n1, n2)
    
    # permutation on counts
    m  <- n1 + n2
    p1 <- sum(n1) / sum(m)
    
    w_perm <- perm_wasserstein_counts(m = m, p1 = p1, n_perm = n_perm)
    
    # one-sided: observed >= permuted
    p_perm <- (1 + sum(w_perm >= w_obs)) / (n_perm + 1)
    
    tibble(
      Geneid      = df_gene$Geneid[1],
      cond1       = p[1],
      cond2       = p[2],
      wasserstein = w_obs,
      p_perm      = p_perm
    )
  })
  
  res <- bind_rows(res_list)
  if (nrow(res) == 0L) return(res)
  
  res
}

## ------------------------------------------------------------------
## 4. Run biological tests across all genes in parallel
## ------------------------------------------------------------------
gene_list <- split(polyA, polyA$Geneid)

# You have 36 cores; use e.g. 32 and leave some free
n_cores <- 32

stats_polyA <- bind_rows(
  mclapply(
    gene_list,
    test_polyA_counts,
    n_perm    = 1000,  # adjust as needed
    min_total = 50,     # or e.g. 500/1000 if you want to filter low coverage
    mc.cores  = n_cores
  )
)

# FDR correction across all gene–pair tests
stats_polyA <- stats_polyA %>%
  mutate(
    p_perm_fdr = p.adjust(p_perm, method = "fdr")
  )


## ------------------------------------------------------------------
## 5. Baseline: technical replicate Wasserstein (same condition, diff '-rep')
## ------------------------------------------------------------------
# For each gene:
#   - group by tech_group (here: condition)
#   - within each tech_group, compute all pairwise sample Wassersteins

compute_w_tech_gene <- function(df_gene) {
  techs <- unique(df_gene$tech_group)
  
  # only keep groups with ≥2 samples
  techs_valid <- techs[sapply(techs, function(tg) {
    sum(df_gene$tech_group == tg) > 1
  })]
  
  if (length(techs_valid) == 0) return(tibble())
  
  res <- map_dfr(techs_valid, function(tg) {
    sub <- df_gene %>% filter(tech_group == tg)
    
    samples <- unique(sub$sample)
    if (length(samples) < 2L) return(tibble())
    
    pairs <- combn(samples, 2, simplify = FALSE)
    
    map_dfr(pairs, function(p) {
      s1 <- p[1]
      s2 <- p[2]
      
      tmp <- df_gene %>% filter(sample %in% c(s1, s2))
      
      lengths <- sort(unique(tmp$polyA))
      
      wide <- tmp %>%
        group_by(sample, polyA) %>%
        summarise(n = sum(n), .groups = "drop") %>%
        tidyr::complete(sample, polyA = lengths, fill = list(n = 0)) %>%
        tidyr::pivot_wider(names_from = sample, values_from = n)
      
      n1 <- as.numeric(wide[[s1]])
      n2 <- as.numeric(wide[[s2]])
      
      tibble(
        Geneid      = df_gene$Geneid[1],
        tech_group  = tg,
        sample1     = s1,
        sample2     = s2,
        wasserstein = wasserstein_counts(n1, n2)
      )
    })
  })
  
  res
}

# Run for all genes (can parallelise, but usually cheaper than the perm test)
tech_w <- bind_rows(
  mclapply(
    gene_list,
    compute_w_tech_gene,
    mc.cores = n_cores
  )
)

## ------------------------------------------------------------------
## 6. Some optional summaries for baseline noise
## ------------------------------------------------------------------

# Global baseline distribution of technical replicate W1
summary(tech_w$wasserstein)
quantile(tech_w$wasserstein, c(0.5, 0.9, 0.95, 0.99), na.rm = TRUE)

# Gene-specific thresholds (e.g. 95% quantile per gene)
tech_w_gene_thresholds <- tech_w %>%
  group_by(Geneid) %>%
  summarise(
    n_pairs  = n(),
    median_w = median(wasserstein),
    p95_w    = quantile(wasserstein, 0.95),
    .groups  = "drop"
  )

stats_with_baseline <- stats_polyA %>%
  left_join(tech_w_gene_thresholds, by = "Geneid") %>% 
  mutate(W_norm = wasserstein / p95_w) %>% 
  filter(W_norm >= 1.5,
         p_perm_fdr <= 0.05)

stats_polyA_pub <- stats_polyA %>%
  left_join(tech_w_gene_thresholds, by = "Geneid") %>% 
  filter(wasserstein > p95_w +1,
         p_perm_fdr <= 0.05) %>% 
  select(Geneid, cond1, cond2, wasserstein, "FDR" = p_perm_fdr) %>% 
  mutate(Geneid = base::gsub("_F", "", Geneid),
         wasserstein = round(wasserstein, 2),
         FDR = signif(FDR, 3),
         cond1 = case_when(
           cond1 == "par" ~ "mRSiC-FLAG -ATc",
           cond1 == "pr0" ~ "rmRSiC-FLAG -ATc",
           cond1 == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           cond1 == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           cond1 == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         ),
         cond2 = case_when(
           cond2 == "par" ~ "mRSiC-FLAG -ATc",
           cond2 == "pr0" ~ "rmRSiC-FLAG -ATc",
           cond2 == "pr1" ~ "rmRSiC-FLAG 1d ATc",
           cond2 == "pr2" ~ "rmRSiC-FLAG 2d ATc",
           cond2 == "pr3" ~ "rmRSiC-FLAG 3d ATc"
         )) %>% 
  write_csv(file = "/run/user/1000/gvfs/smb-share:server=hubiombp12c.cms.hu-berlin.de,share=biomolgenetik/Lab4/Sabrina/HPR4/paper/tables/TableS4_polyAStats")

condition_labels <- c(
  "par" = "mRSiC-FLAG -ATc",
  "pr0" = "rmRSiC-FLAG -ATc",
  "pr1" = "rmRSiC-FLAG 1d ATc",
  "pr2" = "rmRSiC-FLAG 2d ATc",
  "pr3" = "rmRSiC-FLAG 3d ATc"
)

gc()

```


## Antisense Analysis

```{r Normalize Antisense counts}

# Load antisense counts
antisenseCountFiles <- list.files("../../out/counts/mito/shortFeatures/antisense",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

nSample <- length(antisenseCountFiles)

absoluteNormalizators <- base::readRDS("intermediates/absolute_norm_values.RDS")

coldata <- read.csv("../../samples.csv")

antisenseMitoCounts <- read.delim(antisenseCountFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in antisenseCountFiles[2:length(antisenseCountFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  antisenseMitoCounts <- left_join(antisenseMitoCounts, counts, by = "Geneid")
}

cleanAntisenseCounts <- antisenseMitoCounts %>% 
  mutate(Geneid = base::gsub(".rRNA", "", Geneid)) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

# Normalize as before
# mitoCounts: matrix of mito features (rows) x samples (columns), raw counts
mitoCounts <- cleanAntisenseCounts %>%
  mutate(Geneid = paste0(Geneid, "XX")) %>%
  column_to_rownames("Geneid") 

# group: factor of conditions for each sample (length = ncol(mitoCounts))
group <- factor(base::substr(colnames(mitoCounts), 1,3), levels = unique(base::substr(colnames(mitoCounts), 1,3)))

# anchor A_j:   numeric vector of length ncol(mitoCounts)
#               IMPORTANT: A_j must exclude mitochondrial reads; >0 in every sample
Aj <- absoluteNormalizators$value


# 1) Prepare DGEList with offset-only normalization (NO TMM)
Aj <- Aj * (1e6 / median(Aj))       # optional rescale for stability
y  <- DGEList(mitoCounts)

y$samples$lib.size <- Aj
y$samples$group <- group

# 2) Design with all groups (no intercept), pr_0 will be the baseline in contrasts
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

# 3) Dispersion + fit (QL pipeline)
y  <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# 4) Define contrasts: (first) - (pr_0)
CT <- makeContrasts(
  par_pr0 = par - pr0,
  pr0_pr1 = pr1    - pr0,   # logFC > 0 => pr_1 > pr_0
  pr0_pr2 = pr2    - pr0,
  pr0_pr3 = pr3    - pr0,
  levels = design
)

# 5) Run tests and collect results
run_one <- function(fit, contr_name) {
  tt <- glmQLFTest(fit, contrast = CT[, contr_name])
  out <- topTags(tt, n = Inf)$table
  out$contrast <- contr_name
  out
}

res_list <- lapply(colnames(CT), function(nm) run_one(fit, nm))
resAntisense <- do.call(rbind, res_list)

# res has logFC (relative to the anchor), PValue, FDR for each feature/bin and contrast
split(resAntisense, resAntisense$contrast) |>
  lapply(function(df) df[order(df$FDR), ]) -> by_contrast

resAntisenseClean <- resAntisense %>% 
  rownames_to_column("Geneid") %>% 
  mutate(Geneid = sub("XX.*", "", Geneid))


normAntisenseReads <- data.frame(norm_counts <- cpm(y, normalized.lib.sizes = TRUE)) %>%
  rownames_to_column("Geneid") %>%
  mutate(Geneid = sub("XX", "", Geneid)) %>%
  pivot_longer(cols = 1:nSample+1,
               names_to = "sample_name",
               values_to = "antisense") %>% 
  mutate(con = base::substr(sample_name, 1,3)) %>% 
  group_by(Geneid) %>% 
  filter(max(antisense) >= 10) %>% 
  left_join(normReads, by = c("Geneid", "sample_name")) %>% 
  select(Geneid, antisense, sense = value, con = con.x) %>% 
  pivot_longer(cols = c("sense", "antisense"))

for (id in unique(normAntisenseReads$Geneid)){
  plot <- ggplot(filter(normAntisenseReads, Geneid == id),
                 aes(x = con,
                     y = value,
                     color = name)) +
    geom_point() +
    theme_bw() +
    scale_y_log10() +
    labs(title = id,
         y = expression(log[10]("Normalized Reads")),
         x = "Condition")
  
  png(filename = paste0("plots/antisense/", id, ".png"),
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
  plot(plot)
  dev.off()  
  plot(plot)
}

```


```{r save plots, results='hide'}
png(filename = "plots/heatmap_all_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_all)
dev.off()

png(filename = "plots/heatmap_significant_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_sig)
dev.off()

png(filename = "plots/PCA.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(p1)
dev.off()

png(filename = "plots/volcano_d0_d1.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_1)
dev.off()

png(filename = "plots/volcano_d0_d2.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_2)
dev.off()

png(filename = "plots/volcano_d0_d3.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_3)
dev.off()

png(filename = "plots/relativeChanged_FL.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_FL)
dev.off()

png(filename = "plots/relativeChanged_all.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_all)
dev.off()
```
