---
title: "DESEQ_analysis"
author: "Sascha Maschmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

packages <- c("plotly", "tidyverse", "ggrepel", "circlize")

installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)

if (length(to_install) > 0) {
  message("Installing missing packages: ", paste(to_install, collapse = ", "))
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

library(plotly)
library(tidyverse)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(edgeR)

```

## Analysis of mito reads

After finding suitable scaling genes, we use their total read numbers per sample normalize the mito mapped reads.

```{r load data}
countFiles <- list.files("../../out/counts",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

nSample <- length(countFiles)

absoluteNormalizators <- base::readRDS("intermediates/absolute_norm_values.RDS")

coldata <- read.csv("../../samples.csv")

mitoCounts <- read.delim(countFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in countFiles[2:length(countFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  mitoCounts <- left_join(mitoCounts, counts, by = "Geneid")
}


# Filter out all genes with less than 25 reads in all samples
cleanCounts <- mitoCounts %>% 
  filter(if_all(-Geneid, ~ .x >= 25)) %>% 
  # Remove 'rRNA' from Geneids
  mutate(Geneid = base::gsub(".rRNA", "", Geneid)) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  merge(coldata, by = "sample") %>% 
  select(sample = sample_name, Geneid, value) %>% 
  mutate(sample = base::gsub("\\-", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

```

```{r plot raw reads for example genes, fig.asp=9/19, fig.width=12}
# Set example genes throughout the document
example_genes = c("RNA36", "RNA33_F", "RNA2", "RNA30", "RNA18", "SSUD", "SSUE_2", "SSUE_F", "SSUE_1")

example_plot <- cleanCounts %>% 
  filter(Geneid %in% example_genes) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample_name") %>% 
  mutate(sample_name = base::gsub("\\_", "-", sample_name)) %>% 
  merge(coldata, by = "sample_name") 

ggplot(example_plot,
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  theme_bw() +
  scale_y_log10() +
  labs(x = "Condition",
       y = expression(log[10]("Absolute Reads")),
       title = "Raw Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")

```

```{r Normalize}
# mitoCounts: matrix of mito features (rows) x samples (columns), raw counts
mitoCounts <- cleanCounts %>%
  mutate(Geneid = paste0(Geneid, "XX")) %>%
  column_to_rownames("Geneid") 

# group: factor of conditions for each sample (length = ncol(mitoCounts))
group <- factor(base::gsub("_Rep.*", "", colnames(mitoCounts)), levels = unique(base::gsub("_Rep.*", "", colnames(mitoCounts))))

# anchor A_j:   numeric vector of length ncol(mitoCounts)
#               IMPORTANT: A_j must exclude mitochondrial reads; >0 in every sample
Aj <- absoluteNormalizators$value


# 1) Prepare DGEList with offset-only normalization (NO TMM)
Aj <- Aj * (1e6 / median(Aj))       # optional rescale for stability
y  <- DGEList(mitoCounts)

y$samples$lib.size <- Aj
y$samples$group <- group

# 2) Design with all groups (no intercept), pr_0 will be the baseline in contrasts
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

# 3) Dispersion + fit (QL pipeline)
y  <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# 4) Define contrasts: (first) - (pr_0)
CT <- makeContrasts(
  par0_par3 = par_Day_3 - par_Day_0,
  par0_pr0 = pr_Day_0 - par_Day_0,
  pr0_pr1 = pr_Day_1    - pr_Day_0,   # logFC > 0 => pr_1 > pr_0
  pr0_pr2 = pr_Day_2    - pr_Day_0,
  pr0_pr3 = pr_Day_3    - pr_Day_0,
  levels = design
)

# 5) Run tests and collect results
run_one <- function(fit, contr_name) {
  tt <- glmQLFTest(fit, contrast = CT[, contr_name])
  out <- topTags(tt, n = Inf)$table
  out$contrast <- contr_name
  out
}

res_list <- lapply(colnames(CT), function(nm) run_one(fit, nm))
res <- do.call(rbind, res_list)

# res has logFC (relative to the anchor), PValue, FDR for each feature/bin and contrast
split(res, res$contrast) |>
  lapply(function(df) df[order(df$FDR), ]) -> by_contrast

res_clean <- res %>% 
  rownames_to_column("Geneid") %>% 
  mutate(Geneid = sub("XX.*", "", Geneid))
```

```{r plot normalized reads, fig.asp=9/19, fig.width=12}

normReads <- data.frame(norm_counts <- cpm(y, normalized.lib.sizes = TRUE)) %>%
  rownames_to_column("Geneid") %>%
  mutate(Geneid = sub("XX", "", Geneid)) %>%
  pivot_longer(cols = 1:nSample+1,
               names_to = "sample_name",
               values_to = "value") %>%
  mutate(sample_name = base::gsub("\\_", "-", sample_name)) %>% 
  merge(coldata, by = "sample_name") 


ggplot(filter(normReads, Geneid %in% example_genes),
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Condition",
       y = expression(log[10]("Normalized Reads")),
       title = "Normalized Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")

```

The normalization smoothes the differences between replicates, but preserves the overall trend.

```{r PCA}
PCA_Reads <- cpm(y, log = TRUE, normalized.lib.sizes = TRUE)
PCA_reads <- cleanCounts %>% as.matrix()


pca <- prcomp(t(PCA_Reads), scale. = TRUE)
ve <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)


autoplot <- function(pc, col) {
  df <- data.frame(pc$x[,1:2], group = base::gsub("_Rep.*", "", rownames(pc$x)))
  ggplot(df, aes(PC1, PC2, shape = group)) +
    geom_point(size = 2) +
    labs(x = paste0("PC1 (", ve[1], "%)"),
         y = paste0("PC2 (", ve[2], "%)"),
         title = "PCA of Normalized Mito Reads") +
    theme_bw()
}

p1 <- autoplot(pca, "con")
p1

```

```{r Heatmap all Genes, fig.height=9, fig.width=3.5}

lfc_matrix <- res_clean %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  column_to_rownames("Geneid") %>% 
  as.matrix()

col_fun1 = colorRamp2(c(ceiling(max(lfc_matrix, na.rm = TRUE)), 0, floor(min(lfc_matrix, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_all <- Heatmap(lfc_matrix,
        column_title = "LFCs mito rRNA fragments",
        cluster_columns = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC",
        )
plot(heatmap_all)
```

Based on unfiltered log2fold changes, the general trend continues: over all, mito rRNA transcription presumably is dampened by the suppression of HPR4 transcription.

```{r filtered Heatmap, fig.height=7.5, fig.width=3.5}
p_cutoff <- 0.05
fdr_cutoff <- 0.05

lfc_matrix_minimal <- res_clean %>% 
  mutate(logFC = ifelse(PValue <= p_cutoff & FDR <= fdr_cutoff, logFC, NA)) %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  filter(if_any(-Geneid, ~!is.na(.x))) %>% 
  select_if(~sum(!is.na(.)) > 0) %>% 
  column_to_rownames("Geneid") %>% 
  arrange(pr0_pr3) %>%
  as.matrix()

col_fun2 = colorRamp2(c(2, 0, floor(min(lfc_matrix_minimal, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_sig <- Heatmap(lfc_matrix_minimal,
        column_title = "Significantly changed \n mito rRNA fragments",
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC"
        )
plot(heatmap_sig)
```

When filtering based on adjusted p value (\<= 0.05) only decreases in rRNA fragment expression remain.

```{r Volcano static}
units <- read.csv("intermediates/unit_assignment.csv")

volcano <- res_clean %>% 
  select(-logCPM, -`F`) %>% 
  mutate(y_var = -log10(FDR),
         color = ifelse(FDR <= fdr_cutoff & abs(logFC) >= 1, col_fun1(logFC), "grey" ),
         label = ifelse(PValue <= p_cutoff & FDR <= fdr_cutoff, Geneid, ""),
         Geneid2 = Geneid,
         Geneid = base::gsub("_.*", "", Geneid)
  ) %>% 
  left_join(units, by = "Geneid") %>% 
  mutate(Geneid = Geneid2,
         unit = case_when(!is.na(unit) ~ unit,
                          grepl("LSU", Geneid) ~ "LSU",
                          grepl("SSU", Geneid) ~ "SSU",
                          is.na(unit) ~ "unassigned"
                          )) %>% 
  select(-Geneid2)

# Plot volcano plots
volcano_0_1 <- ggplot(data = filter(volcano, contrast == "pr0_pr1"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 1") +
  # Set theme
  theme_bw() 

# Plot volcano plots
volcano_0_2 <- ggplot(data = filter(volcano, contrast == "pr0_pr2"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 2") +
  # Set theme
  theme_bw() 

volcano_0_3 <- ggplot(data = filter(volcano, contrast == "pr0_pr3"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 3") +
  # Set theme
  theme_bw() 

plot(volcano_0_1)
plot(volcano_0_2)
plot(volcano_0_3)

```

For day 3 comparisons, the plot was made interactive, since the points are too dense for labels. Hover the Points to see the gene.

```{r Volcano interactive}
# Due to the dense Distribution of samples, this plot is created interactively with plotly
volc1 <- ggplot(data = filter(volcano, contrast == "pr0_pr3"),
       aes(x = logFC,
           y = y_var,
           ID = Geneid)) +
  # Plot LFC against FDR
  geom_point(aes(color = color)) +
  scale_color_identity() +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
           linetype = "dashed",
           color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 3") +
  # Set theme
  theme(legend.position = "none") +
  theme_bw() 

suppressWarnings(ggplotly(hide_legend(volc1), tooltip = c("ID")))


```

### Relative number of significantly changed rRNAs

Below two variants of relative number of significantly changed rRNA are plotted, one only respects the full length transcripts, the other all annotated transcripts variants.

```{r changed rRNAs}
all_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue")) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = nrow(cleanCounts),
         relativeValues = n/total*100)

relativeChanged_all <-
  ggplot(all_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on All Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))


total_full <- cleanCounts %>% 
  filter(!grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  summarise(n()) %>% 
  as.integer()

fullLength_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff,
         !grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue"),
         Geneid = base::gsub("_F", "", Geneid)) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = total_full,
         relativeValues = n/total*100)

relativeChanged_FL <- 
  ggplot(fullLength_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on Full Length Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))

relativeChanged_all
relativeChanged_FL
```


```{r save plots, results='hide'}
png(filename = "plots/heatmap_all_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_all)
dev.off()

png(filename = "plots/heatmap_significant_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_sig)
dev.off()

png(filename = "plots/PCA.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(p1)
dev.off()

png(filename = "plots/volcano_d0_d1.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_1)
dev.off()

png(filename = "plots/volcano_d0_d2.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_2)
dev.off()

png(filename = "plots/volcano_d0_d3.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_3)
dev.off()

png(filename = "plots/relativeChanged_FL.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_FL)
dev.off()

png(filename = "plots/relativeChanged_all.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_all)
dev.off()
```
