---
title: "DESEQ_analysis"
author: "Sascha Maschmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

packages <- c("plotly", "tidyverse", "ggrepel", "circlize")

installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)

if (length(to_install) > 0) {
  message("Installing missing packages: ", paste(to_install, collapse = ", "))
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

library(plotly)
library(tidyverse)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(edgeR)

```

# Analysis of mito reads

## Mapping percentages
Investigation of relative amounts of reads mapping to the pseudo-genome.  
Decreases with treatment.

```{r plot mapping percentages, fig.asp=9/19, fig.width=12}
mapping_perc <- read.table("../../out/mapping_percentages.txt", header = T) %>% 
  mutate(con = base::substr(sample, 1,3))

ggplot(mapping_perc,
       aes(x = con,
           y = percent_mapped)) +
  geom_point() +
  theme_bw() +
  labs(x = "Condition",
       y = "Fraction of Total Reads Mapped to Mito Genome [%]",
       title = "Mito Maping Percentages")
```

## Normalisation
Using A set of stable genes encoded in the nuclear or apicoplast genome, mito mapped reads are normalized.

```{r load data}
countFiles <- list.files("../../out/counts/mito/shortFeatures/sense",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

nSample <- length(countFiles)

absoluteNormalizators <- base::readRDS("intermediates/absolute_norm_values.RDS")

coldata <- read.csv("../../samples.csv")

mitoCounts <- read.delim(countFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in countFiles[2:length(countFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  mitoCounts <- left_join(mitoCounts, counts, by = "Geneid")
}

longFiles <- list.files("../../out/counts/mito/longFeatures",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

# Filter out all genes with less than 20 reads in all samples
cleanShortCounts <- mitoCounts %>% 
  filter(if_all(-Geneid, ~ .x >= 20)) %>% 
  # Remove 'rRNA' from Geneids
  mutate(Geneid = base::gsub(".rRNA", "", Geneid)) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

longCounts <- read.delim(longFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in longFiles[2:length(longFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  longCounts <- left_join(longCounts, counts, by = "Geneid")
}

cleanLongCounts <- longCounts %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

cleanCounts <- rbind(cleanShortCounts, cleanLongCounts)
```

```{r plot example genes, fig.asp=9/19, fig.width=12}
example_genes = c("SSUB", "RNA10_F", "RNA10_1", "RNA10_2", "RNA33_F", "RNA33_1", "RNA33_2")


example_plot <- cleanCounts %>% 
  filter(Geneid %in% example_genes) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample_name") %>% 
  mutate(con = base::substr(sample_name, 1,3))

ggplot(example_plot,
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  theme_bw() +
  scale_y_log10() +
  labs(x = "Condition",
       y = expression(log[10]("Absolute Reads")),
       title = "Raw Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")


```
```{r Normalize}
# mitoCounts: matrix of mito features (rows) x samples (columns), raw counts
mitoCounts <- cleanCounts %>%
  mutate(Geneid = paste0(Geneid, "XX")) %>%
  column_to_rownames("Geneid") 

# group: factor of conditions for each sample (length = ncol(mitoCounts))
group <- factor(base::substr(colnames(mitoCounts), 1,3), levels = unique(base::substr(colnames(mitoCounts), 1,3)))

# anchor A_j:   numeric vector of length ncol(mitoCounts)
#               IMPORTANT: A_j must exclude mitochondrial reads; >0 in every sample
Aj <- absoluteNormalizators$value


# 1) Prepare DGEList with offset-only normalization (NO TMM)
Aj <- Aj * (1e6 / median(Aj))       # optional rescale for stability
y  <- DGEList(mitoCounts)

y$samples$lib.size <- Aj
y$samples$group <- group

# 2) Design with all groups (no intercept), pr_0 will be the baseline in contrasts
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

# 3) Dispersion + fit (QL pipeline)
y  <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# 4) Define contrasts: (first) - (pr_0)
CT <- makeContrasts(
  par_pr0 = par - pr0,
  pr0_pr1 = pr1    - pr0,   # logFC > 0 => pr_1 > pr_0
  pr0_pr2 = pr2    - pr0,
  pr0_pr3 = pr3    - pr0,
  levels = design
)

# 5) Run tests and collect results
run_one <- function(fit, contr_name) {
  tt <- glmQLFTest(fit, contrast = CT[, contr_name])
  out <- topTags(tt, n = Inf)$table
  out$contrast <- contr_name
  out
}

res_list <- lapply(colnames(CT), function(nm) run_one(fit, nm))
res <- do.call(rbind, res_list)

# res has logFC (relative to the anchor), PValue, FDR for each feature/bin and contrast
split(res, res$contrast) |>
  lapply(function(df) df[order(df$FDR), ]) -> by_contrast

res_clean <- res %>% 
  rownames_to_column("Geneid") %>% 
  mutate(Geneid = sub("XX.*", "", Geneid))
```

```{r plot normalized reads, fig.asp=9/19, fig.width=12}

normReads <- data.frame(norm_counts <- cpm(y, normalized.lib.sizes = TRUE)) %>%
  rownames_to_column("Geneid") %>%
  mutate(Geneid = sub("XX", "", Geneid)) %>%
  pivot_longer(cols = 1:nSample+1,
               names_to = "sample_name",
               values_to = "value") %>% 
  mutate(con = base::substr(sample_name, 1,3))


ggplot(filter(normReads, Geneid %in% example_genes),
       aes(x = con,
           y = value,
           color = Geneid)) +
  geom_point(show.legend = FALSE) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Condition",
       y = expression(log[10]("Normalized Reads")),
       title = "Normalized Reads per Condition") +
  facet_wrap(~Geneid,
             scales = "free_y")

```
```{r PCA, fig.asp=9/19, fig.width=12}
PCA_Reads <- cpm(y, log = TRUE, normalized.lib.sizes = TRUE)
PCA_reads <- cleanCounts %>% as.matrix()


pca <- prcomp(t(PCA_Reads), scale. = TRUE)
ve <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
pc <- pca

autoplot <- function(pc, col) {
  df <- data.frame(pc$x[,1:2], group = base::substr(rownames(pc$x), 1,3))
  ggplot(df, aes(PC1, PC2, shape = group)) +
    geom_point(size = 2) +
    labs(x = paste0("PC1 (", ve[1], "%)"),
         y = paste0("PC2 (", ve[2], "%)"),
         title = "PCA of Normalized Mito Reads") +
    theme_bw()
}

p1 <- autoplot(pca, "con")
p1

```

## DEG Analysis
Based on the normalized reads, log2fold changes and statistical significance are calculated.  
Below results are shown in various ways.

```{r Heatmap all Genes, fig.height=9, fig.width=3.5}

lfc_matrix <- res_clean %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  column_to_rownames("Geneid") %>% 
  as.matrix()

col_fun1 = colorRamp2(c(ceiling(max(lfc_matrix, na.rm = TRUE)), 0, floor(min(lfc_matrix, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_all <- Heatmap(lfc_matrix,
        column_title = "LFCs mito rRNA fragments",
        cluster_columns = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC",
        )
plot(heatmap_all)
```

Based on unfiltered log2fold changes, the general trend continues: over all, mito rRNA transcription presumably is dampened by the suppression of HPR4 transcription.

```{r filtered Heatmap, fig.height=7.5, fig.width=3.5}
p_cutoff <- 0.05
fdr_cutoff <- 0.05

lfc_matrix_minimal <- res_clean %>% 
  mutate(logFC = ifelse(PValue <= p_cutoff & FDR <= fdr_cutoff, logFC, NA)) %>% 
  select(Geneid, logFC, contrast) %>% 
  pivot_wider(id_cols = Geneid,
              names_from = contrast,
              values_from = logFC) %>% 
  filter(if_any(-Geneid, ~!is.na(.x))) %>% 
  select_if(~sum(!is.na(.)) > 0) %>% 
  column_to_rownames("Geneid") %>% 
  arrange(pr0_pr2, pr0_pr3) %>%
  as.matrix()

col_fun2 = colorRamp2(c(2, 0, floor(min(lfc_matrix_minimal, na.rm = TRUE))),
                 c("red", "white", "blue"))
heatmap_sig <- Heatmap(lfc_matrix_minimal,
        column_title = "Significantly changed \n mito rRNA fragments",
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        col = col_fun1,
        na_col = "grey20",
        name = "LFC"
        )
plot(heatmap_sig)
```

When filtering based on adjusted p value (\<= 0.05) only decreases in rRNA fragment expression remain.

```{r Volcano static, fig.asp=9/19, fig.width=12}
units <- read.csv("intermediates/unit_assignment.csv")

volcano <- res_clean %>% 
  select(-logCPM, -`F`) %>% 
  mutate(y_var = -log10(FDR),
         color = ifelse(FDR <= fdr_cutoff & abs(logFC) >= 1, col_fun1(logFC), "grey" ),
         label = ifelse(PValue <= p_cutoff & FDR <= fdr_cutoff, Geneid, ""),
         Geneid2 = Geneid,
         Geneid = base::gsub("_.*", "", Geneid)
  ) %>% 
  left_join(units, by = "Geneid") %>% 
  mutate(Geneid = Geneid2,
         unit = case_when(!is.na(unit) ~ unit,
                          grepl("LSU", Geneid) ~ "LSU",
                          grepl("SSU", Geneid) ~ "SSU",
                          is.na(unit) ~ "unassigned"
                          )) %>% 
  select(-Geneid2)

# Plot volcano plots
volcano_0_1 <- ggplot(data = filter(volcano, contrast == "pr0_pr1"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 1") +
  # Set theme
  theme_bw() 

# Plot volcano plots
volcano_0_2 <- ggplot(data = filter(volcano, contrast == "pr0_pr2"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 2") +
  # Set theme
  theme_bw() 

volcano_0_3 <- ggplot(data = filter(volcano, contrast == "pr0_pr3"),
       aes(x = logFC,
           y = y_var)) +
  # Plot LFC against FDR
  geom_point(aes(color = color,
                 shape = unit),
             show.legend = TRUE) +
  scale_color_identity() +
  # Add labels
  geom_text_repel(aes(label = label),
                  segment.color = 'black',
                  min.segment.length = .1,
                  max.overlaps = 20,
                  size = 3,
                  force = 8,
                  force_pull = 0.2) +
  # Add cutoff indicators
  geom_hline(yintercept = -log10(fdr_cutoff),
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = -1,
             linetype = "dashed",
             color = "grey40") +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "grey40") +
  # Adjust titles
  labs(x = expression(log[2]("fold change")),
       y = expression(-log[10](padj)),
       title = "Day 0 vs Day 3") +
  # Set theme
  theme_bw() 

plot(volcano_0_1)
plot(volcano_0_2)
plot(volcano_0_3)

```

## Quantification of Subunits affected
The table below highlights the ratio of large and small ribosome subunit fragments.

```{r quantify unit effect}

unitEffect <- volcano %>% 
  select(Geneid, FDR, logFC, contrast, unit) %>% 
  filter(FDR <= fdr_cutoff) %>% 
  mutate(expressed = ifelse(sign(logFC) == -1, "down", "up")) %>% 
  group_by(contrast, expressed) %>% 
  mutate(totalDE = n()) %>% 
  ungroup() %>% 
  group_by(contrast, unit, expressed) %>% 
  summarise(relative = n() / totalDE) %>% 
  distinct()

print(unitEffect)
```

## Compare Fragments per RNA 'identity'
Grey Log2FCs are not significant.

```{r multi fragment cases, fig.asp=9/19, fig.width=12}

deMultiFrag <- res_clean %>% 
  mutate(Geneid2 = Geneid,
         Geneid = base::gsub("_.*", "", Geneid)
  ) %>% 
  select(Geneid, Geneid2, logFC, contrast, FDR) %>% 
  group_by(Geneid)  %>% 
  group_split()

test <- deMultiFrag[[17]] %>% 
  mutate(Geneid2 = ifelse(grepl("_", Geneid2), sub("^[^_]*_", "", Geneid2), "F")) %>% 
  pivot_wider(values_from = c(logFC, FDR), names_from = Geneid2)

untangler <- function(df){
  df_untangled <- df %>% 
    mutate(Geneid2 = ifelse(grepl("_", Geneid2), sub("^[^_]*_", "", Geneid2), "F")) %>% 
    pivot_wider(values_from = c(logFC, FDR), names_from = Geneid2)
  return(df_untangled)
}

fragTable <- lapply(deMultiFrag, untangler) %>%
  bind_rows()

trueFragmented <- fragTable %>% 
  filter(!is.na(logFC_1)) %>% 
  pivot_longer(cols = 3:8,
               names_to = c("stat", "fragment"),
               names_sep = "_") %>% 
  pivot_wider(names_from = "stat",
              values_from = "value") %>% 
  mutate(color = ifelse(FDR <= fdr_cutoff, "black", "grey"))

for (id in unique(trueFragmented$Geneid)){
  plot <- ggplot(
    data = filter(trueFragmented, Geneid == id),
    aes(x = contrast,
        y = logFC, 
        shape = fragment,
        color = color)
    ) +
    scale_color_identity() +
    geom_point(size = 2) +
    theme_bw() +
    labs(x = "",
         y = 'LFC',
         title = id)
  
  plot(plot)
}
```


## Relative number of significantly changed rRNAs

Below two variants of relative number of significantly changed rRNA are plotted, one only respects the full length transcripts, the other all annotated transcripts variants.

```{r changed rRNAs, fig.asp=9/19, fig.width=12}
all_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue")) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = nrow(cleanCounts),
         relativeValues = n/total*100)

relativeChanged_all <-
  ggplot(all_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on All Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))


total_full <- cleanCounts %>% 
  filter(!grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  summarise(n()) %>% 
  as.integer()

fullLength_changed <- res_clean %>% 
  filter(FDR <= fdr_cutoff,
         !grepl("_1", Geneid),
         !grepl("_2", Geneid)) %>% 
  mutate(expressed = ifelse(logFC > 0, "firebrick1", "dodgerblue"),
         Geneid = base::gsub("_F", "", Geneid)) %>% 
  group_by(contrast, expressed) %>% 
  summarise(n = n()) %>% 
  mutate(n = ifelse(expressed == "dodgerblue", n*-1, n),
         total = total_full,
         relativeValues = n/total*100)

relativeChanged_FL <- 
  ggplot(fullLength_changed,
       aes(x = contrast,
           y = relativeValues,
           label = abs(n),
           fill = expressed)) +
  geom_bar(stat = "identity") +
  geom_label(aes(y = relativeValues + 1.25 * sign(relativeValues))) +
  scale_fill_identity() +
  theme_bw() +
  labs(x = "Comparison",
       y = "Relative number of rRNAs [%]",
       title = "Number of Significantly Changed rRNAs Based on Full Length Fragments",
       caption = paste("Labels are absolute numbers of rRNA fragments with adjusted p-values below", fdr_cutoff))

relativeChanged_all
relativeChanged_FL
```

## PolyA Tail plots

```{r polya tail plots, fig.asp=9/19, fig.width=12}
polyA_files <- list.files("../../out/polyA",
                          pattern = "*_polyA_lengths.tsv$",
                          full.names = T)
polyA <- data_frame()
for (f in polyA_files){
  ph <- read.delim(f, header = F) %>% 
    select(Geneid = V1, polyA = V2, n = V3) %>% 
    mutate(sample = f,
           sample = base::gsub("../../out/polyA/", "", sample),
           sample = base::gsub("_polyA_lengths.tsv", "", sample))
  
  polyA <- rbind(polyA, ph)
}

polyA_summarised <- polyA %>% 
  group_by(Geneid, sample) %>% 
  arrange(polyA) %>% 
  mutate(n_rel = n / sum(n),
         cumsum = cumsum(n_rel),
         n_total = sum(n)) %>% 
  filter(cumsum <= 0.99,
         n_total >= 25) %>% 
  ungroup() %>% 
  mutate(condition = base::substr(sample, 1,3),
         Geneid = base::gsub("/", "-", Geneid)) %>% 
  group_by(Geneid, polyA, condition) %>% 
  summarise(n_mean = mean(n_rel),
            stdDev = sd(n_rel))

for (id in unique(polyA_summarised$Geneid)){
  plot <-
    ggplot(filter(polyA_summarised, Geneid == id),
                 aes(x = polyA,
                     y = n_mean,
                     color = condition,
                     fill = condition,
                     ymin = n_mean - stdDev,
                     ymax = n_mean + stdDev)) +
    geom_linerange(aes()) +
    geom_line() +
    theme_bw() +
    labs(x = "polyA-tail length",
         y = "mean relative frequency",
         title = id)
  
  png(filename = paste0("plots/polyA/", id, ".png"),
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
  plot(plot)
  dev.off()  
  plot(plot)
}
```

## Antisense Analysis

```{r Normalize Antisense counts}

# Load antisense counts
antisenseCountFiles <- list.files("../../out/counts/mito/shortFeatures/antisense",
                         pattern = "*mito_featureCounts.txt$",
                         full.names = TRUE)

nSample <- length(antisenseCountFiles)

absoluteNormalizators <- base::readRDS("intermediates/absolute_norm_values.RDS")

coldata <- read.csv("../../samples.csv")

antisenseMitoCounts <- read.delim(antisenseCountFiles[1], header = T, skip = 1) %>% 
  select(Geneid, last_col())
for (file in antisenseCountFiles[2:length(antisenseCountFiles)]){
  counts <- read.delim(file, header = T, skip = 1) %>% 
    select(Geneid, last_col())
  antisenseMitoCounts <- left_join(antisenseMitoCounts, counts, by = "Geneid")
}

cleanAntisenseCounts <- antisenseMitoCounts %>% 
  mutate(Geneid = base::gsub(".rRNA", "", Geneid)) %>% 
  pivot_longer(cols = 1:nSample+1,names_to = "sample") %>% 
  mutate(sample = base::gsub("out.filtered.", "", sample),
         sample = base::gsub("_mito_filtered.bam", "", sample),
         sample = base::gsub("\\.", "_", sample)) %>% 
  pivot_wider(names_from = sample,
              values_from = value)

# Normalize as before
# mitoCounts: matrix of mito features (rows) x samples (columns), raw counts
mitoCounts <- cleanAntisenseCounts %>%
  mutate(Geneid = paste0(Geneid, "XX")) %>%
  column_to_rownames("Geneid") 

# group: factor of conditions for each sample (length = ncol(mitoCounts))
group <- factor(base::substr(colnames(mitoCounts), 1,3), levels = unique(base::substr(colnames(mitoCounts), 1,3)))

# anchor A_j:   numeric vector of length ncol(mitoCounts)
#               IMPORTANT: A_j must exclude mitochondrial reads; >0 in every sample
Aj <- absoluteNormalizators$value


# 1) Prepare DGEList with offset-only normalization (NO TMM)
Aj <- Aj * (1e6 / median(Aj))       # optional rescale for stability
y  <- DGEList(mitoCounts)

y$samples$lib.size <- Aj
y$samples$group <- group

# 2) Design with all groups (no intercept), pr_0 will be the baseline in contrasts
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)

# 3) Dispersion + fit (QL pipeline)
y  <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

# 4) Define contrasts: (first) - (pr_0)
CT <- makeContrasts(
  par_pr0 = par - pr0,
  pr0_pr1 = pr1    - pr0,   # logFC > 0 => pr_1 > pr_0
  pr0_pr2 = pr2    - pr0,
  pr0_pr3 = pr3    - pr0,
  levels = design
)

# 5) Run tests and collect results
run_one <- function(fit, contr_name) {
  tt <- glmQLFTest(fit, contrast = CT[, contr_name])
  out <- topTags(tt, n = Inf)$table
  out$contrast <- contr_name
  out
}

res_list <- lapply(colnames(CT), function(nm) run_one(fit, nm))
resAntisense <- do.call(rbind, res_list)

# res has logFC (relative to the anchor), PValue, FDR for each feature/bin and contrast
split(resAntisense, resAntisense$contrast) |>
  lapply(function(df) df[order(df$FDR), ]) -> by_contrast

resAntisenseClean <- resAntisense %>% 
  rownames_to_column("Geneid") %>% 
  mutate(Geneid = sub("XX.*", "", Geneid))


normAntisenseReads <- data.frame(norm_counts <- cpm(y, normalized.lib.sizes = TRUE)) %>%
  rownames_to_column("Geneid") %>%
  mutate(Geneid = sub("XX", "", Geneid)) %>%
  pivot_longer(cols = 1:nSample+1,
               names_to = "sample_name",
               values_to = "antisense") %>% 
  mutate(con = base::substr(sample_name, 1,3)) %>% 
  group_by(Geneid) %>% 
  filter(max(antisense) >= 10) %>% 
  left_join(normReads, by = c("Geneid", "sample_name")) %>% 
  select(Geneid, antisense, sense = value, con = con.x) %>% 
  pivot_longer(cols = c("sense", "antisense"))

for (id in unique(normAntisenseReads$Geneid)){
  plot <- ggplot(filter(normAntisenseReads, Geneid == id),
                 aes(x = con,
                     y = value,
                     color = name)) +
    geom_point() +
    theme_bw() +
    scale_y_log10() +
    labs(title = id,
         y = expression(log[10]("Normalized Reads")),
         x = "Condition")
  
  png(filename = paste0("plots/antisense/", id, ".png"),
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
  plot(plot)
  dev.off()  
  plot(plot)
}

```


```{r save plots, results='hide'}
png(filename = "plots/heatmap_all_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_all)
dev.off()

png(filename = "plots/heatmap_significant_LFC.png",
       width = 10,
       height = 25,
       units = "cm",
    res = 400)
plot(heatmap_sig)
dev.off()

png(filename = "plots/PCA.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(p1)
dev.off()

png(filename = "plots/volcano_d0_d1.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_1)
dev.off()

png(filename = "plots/volcano_d0_d2.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_2)
dev.off()

png(filename = "plots/volcano_d0_d3.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(volcano_0_3)
dev.off()

png(filename = "plots/relativeChanged_FL.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_FL)
dev.off()

png(filename = "plots/relativeChanged_all.png",
       width = 16,
       height = 9,
       units = "cm",
    res = 400)
plot(relativeChanged_all)
dev.off()
```
